---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

* * *

## Part 0: Introduction

It's been more than six years since President Obama signed the Affordable Care Act into law. The administration credits Obamacare -- as it is colloquially known -- for lowering the number of uninsured residents: down 11 million since its passage, according to a March 2015 report by the Centers for Disease Control and Prevention.

The uninsured rate varies between states, from 17% in Texas to 5% in Massachusetts, according to the Kaiser Family Foundation. Given this range of uninsured residents, **is there an association between health care coverage and general health-related characteristics?**

I'll try to answer this question by looking at the [Behavioral Risk Factor Surveillance System](http://www.cdc.gov/brfss), a yearly telephone survey that collects data on health-related risk behaviors.

* * *

## Part 1: Data

The population of interest is non-institutionalized US residents 18 years or older. To perform the survey, BRFSS researchers use Random Digit Dialing to call thousands of adults -- 506,000 in 2014 -- in all 50 states, DC, and three US territories.

The study might suffer from **coverage bias** because not all US residents own a landline or cellphone. This might mean the study underrepresents hard-to-reach groups.

As an observational study, the BRFSS is conducted by random sampling, which means the survey results can be generalized to the population (assuming there is no bias and the sample is representative of the population.) Since there is no random assignment, the survey cannot be used to establish causality relationships.

* * *

## Part 2: Research questions

**Research question 1:** Is health insurance tied to better health? I'll answer that by looking at the association between health care coverage and general health.

**Research question 2:** If you have insurance, is it more likely you'll be able to afford a doctor visit? Also, which states have the highest proportion of insured residents unable to see a doctor due to cost? I'll answer these by looking at the association between health care coverage and not being able to see a doctor due to cost.

**Research question 3:**

* * *

## Part 3: Exploratory data analysis

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(gmodels)
setwd("~/Desktop/week5")
```

### Load data

```{r load-data}
load('brfss2013.RData')
```

* * *

**Research question 1:** Is general health associated to a lack of health care coverage? Let's look at a contingency table of general health (*genhlth*) and health care coverage (for respondents aged 18-64 with any form of coverage) (*X_hcvu651*):

```{r}

health_coverage.table <- prop.table(table(brfss2013$genhlth, brfss2013$X_hcvu651), 2)
health_coverage.table

```

The proportions vary within general health, which provides evidence that these two variables are associated. We can also look at a couple graphs - first, a histogram:

```{r}

# Convert to data frame for plotting
health_coverage.data <- as.data.frame(health_coverage.table) %>%
	rename(Health = Var1, Coverage = Var2, Proportion = Freq)

ggplot(health_coverage.data, aes(Health, Proportion, fill=Coverage)) +
	geom_bar(stat='identity', position='dodge') +
	labs(
		title='Distribution of general health by health care coverage',
		subtitle='Respondents are aged 18-64 who have any form of health care coverage'
	)

```

Next, a mosaic-like bar plot:

```{r}

ggplot(health_coverage.data, aes(Coverage, Proportion, fill=Health)) +
	geom_bar(stat='identity') +
	labs(
		title='Proportions of general health by health care coverage',
		subtitle='Respondents are aged 18-64 who have any form of health care coverage'
	)

```

In both these graphs we see that the general health proportions vary by coverage, which provides visual evidence that these two variables are associated.

* * *

**Research question 2:** Is there an association between health care coverage and not being able to see a doctor because of cost? Here's another contingency table, this time of health care coverage (for respondents aged 18-64 with any form of coverage) (*X_hcvu651*) and not being able to see the doctor due to cost (*medcost*): 

```{r}

coverage_cost.table <- prop.table(table(brfss2013$X_hcvu651, brfss2013$medcost), 1)
coverage_cost.table

```

Of those residents with insurance, about 10% say they couldn't see the doctor due to cost. This tells me the two variables are associated. We can reach the same conclusion graphically -- here's a mosaic plot:

```{r}

# Convert to data frame for plotting
as.data.frame(coverage_cost.table) %>%
	rename(Coverage = Var1, Doctor.too.costly = Var2, Proportion = Freq) %>%
	ggplot(aes(Coverage, Proportion, fill=Doctor.too.costly)) +
	geom_bar(stat='identity') +
	labs(
		title='Inability to see doctor due to cost by health care coverage',
		subtitle='Respondents are aged 18-64 who have any form of health care coverage'
	)

```

* * * 

Next I'll look at the states with the highest proportion of insured residents unable to see the doctor due to cost.

```{r}

cost.states <- brfss2013 %>%
	select(Coverage = X_hcvu651, Cost = medcost, State = X_state) %>%
	filter(Coverage == 'Have health care coverage') %>%
	select(-Coverage) %>%
	na.omit %>%
	group_by(State, Cost) %>%
	summarise(n = n()) %>%
	mutate(Doctor.too.costly = n / sum(n)) %>%
	filter(Cost == 'Yes') %>%
	select(State, Doctor.too.costly) %>%
	arrange(-Doctor.too.costly)

cost.states

```

The first two are territories. The rest of the top ten entries are states with varying demographics: New Mexico has a large share of hispanics, Mississippi is the poorest state, Florida has a high percentage of retirees... I'm curious as to the driving force behind these proportions. Is it correlated to the state's uninsured rate? 

```{r}

brfss2013 %>%
	select(Coverage = X_hcvu651, State = X_state) %>%
	na.omit %>%
	group_by(State, Coverage) %>%
	summarise(n = n()) %>%
	mutate(Uninsured = n / sum(n)) %>%
	filter(Coverage == 'Do not have health care coverage') %>%
	select(State, Uninsured) %>%
	inner_join(cost.states, by=c('State')) %>%
	ggplot(aes(Doctor.too.costly, Uninsured)) +
	geom_point() +
	scale_x_continuous(labels = scales::percent) + scale_y_continuous(labels = scales::percent) +
	xlab('Residents unable to see doctor due to cost') +
	ylab('Uninsured') +
	labs(
		title = 'Uninsured rate and inability to see doctor due to cost, by state',
		subtitle='Respondents are aged 18-64 who do not have any form of health care coverage'
	)

```

Visually the correlation seems pretty high - as the uninsured rate increases, so does the state's proportion of residents unable to see the doctor due to cost. However, I think maybe this is a flawed comparison: if a state has a high number of uninsured residents, then most likely a large proportion of them will not be able to see the doctor, unless the median income is really high -- correct? I'm not 100% sure on the validity of this graph, so I'll keep it here -- it is visually compelling.

```{r}

brfss2013 %>%
	select(Coverage = X_hcvu651, Age = X_age80) %>%
	filter(Age > 17, Age < 65) %>%
	na.omit %>%
	group_by(Coverage, Age) %>%
	summarise(n = n()) %>%
  mutate(Proportion = n / sum(n)) %>%
	ggplot(aes(Age, Proportion, color=Coverage)) +
	geom_step()

```

```{r}

```















