---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded your data file will be called `brfss2013`. **Delete this note before you submit your work.** 

```{r load-data}
#load("brfss2013.RData")
```



* * *

## Part 1: Data

The [Behavioral Risk Factor Surveillance System](http://www.cdc.gov/brfss) is a yearly telephone survey that collects data on health-related risk behaviors in all 50 states, DC, and three US territories.

As an observational study, the BRFSS is conducted by random sampling. This means that the survey results can be generalized to the population, but since there is no random assignment, there can be no causality.

* * *

## Part 2: Research questions

**Research question 1:**

**Research question 2:**

**Research question 3:**


* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" button (green button with orange arrow) above. Make sure that your code is visible in the project you submit. Delete this note before you submit your work.

**Research question 1:**

```{r}

```



**Research question 2:**

```{r}

```



**Research question 3:**

```{r}

```

